Enhancing Adversarial Robustness for Deep Metric Learning

### 2021-10-26 robdml.tex
tex

@@

Given an image triplet ($\mA$, $\mP$, $\mN$) sampled with a certain sampling
strategy (\eg, semi-hard~\cite{facenet}) within a mini-batch, we define its
\emph{hardness} as a scalar:
%
\begin{equation}
%
H(\mA,\mP,\mN)=d_\phi(\mA,\mP)-d_\phi(\mA,\mN),
%
\end{equation}
%
which is an internal part of the triplet loss.
%
For convenience we call this triplet as ``\emph{source} triplet'', and the
corresponding $H(\cdots)$ value as ``\emph{source} hardness'', denoted as
$H_\mathsf{S}$.

Then, \emph{Hardness Manipulation} (HM) aims to increase the \emph{source}
hardness $H_\mathsf{S}$ into a specified ``\emph{destination} hardness''
$H_\mathsf{D}$, by finding adversarial examples of the source triplet $(\mA,
\mP, \mN)$, \ie, $(\mA + {\vr}_a, \mP + {\vr}_p, \mN + {\vr}_n)$, where
$({\vr}_a, {\vr}_p, {\vr}_n)$ are the adversarial perturbations.
%
If we denote the hardness of the adversarially perturbed \emph{source} triplet
as $\tilde{H}_\mathsf{S}$, the HM can be implemented as follows:
%
\begin{equation}
	%
	\hat{\vr}_a, \hat{\vr}_p, \hat{\vr}_n = \argmin_{\vr_a,\vr_p,\vr_n}
	\big\|\max(0, H_\mathsf{D} - \tilde{H}_\mathsf{S} ) \big\|_2^2.
	%
	\label{eq:hm}
	%
\end{equation}
%
The $\max(0,\cdot)$ part in \cref{eq:hm} truncates the gradient when
$H_\mathsf{S}>H_\mathsf{D}$, automatically stopping the optimization, because
$H_\mathsf{S}$ is not desired to be reduced once being greater than
$H_\mathsf{D}$.
%
The \cref{eq:hm} is written in the L-$2$ norm form instead of the standard Mean Squared
Error because HM can be directly extended into vector form for a
mini-batch.
%
The optimization problem can be solved by Projected Gradient
Descent (PGD)~\cite{madry}.
%
And the resulting adversarial examples are used for adversarially training the
DML model:
%
$L_\text{T}(\phi(\mA+\hat{\vr}_a), \phi(\mP+\hat{\vr}_p),
\phi(\mN+\hat{\vr}_n))$.
%
We abbreviate the adversarial training with adversarial examples created through
this way as $\text{HM}[H_\mathsf{S},H_\mathsf{D}]$ in the following text.

@@

Instead of directly maximizing the source hardness, we propose to stop the
maximization process when the adversarial examples reach a certain level of
hardness in order to avoid model collapse.

\oo{uses gradient for maximization. Maxim is $HM(H_\mathsf{S}, 2)$.}

In the traditional min-max adversarial training setting~\cite{madry}, the inner
``max'' problem should maximize $H(X_a,X_p^S,X_n^S)$ and hence maximize the triplet
loss.

In this paper, we argue that the hardness of triplet matters in adversarial
training for DML.
%
The model does not collapse with EST or ACT because the expectation of hardness,
\ie, $E[H]$ will be close to zero.
%
While the variance $\text{Var}[H]$ for EST is higher, and that for ACT is lower.

\cref{fig:hm}
%
Note, it is not expected to lower the hardness during the attack, namely
$E[H_\mathsf{D}]$ should be greater or equal to $E[H_\mathsf{S}]$.
%
%

%
Based on the same anchor image $X_a$, we sample another positive image $X_p^D$
and negative image $X_n^D$ with a different triplet sampling strategy, and
we call its hardness as ``destination hardness'', denoted as $H_\mathsf{D}$.

Note, the design of HM is flexible, and not coupled with specific sampling
strategy.
%
In other words, any existing or future triplet sampling strategy can
be adopted in HM.
%
or even a pseudo hardness function $g(\cdot)$ as long as we have a strong
prior knowledge on what the destination hardness should be.

\oo{extend: fat}
training process of a regular DML model, but existing defense methods are
incompatible to acceleration methods like Free Adversarial
Training~\cite{freeat}.

\oo{[4. Free Adversarial Metric Learining]}
Extension of FAT to adversarial deep metric learning.

%
Assume there are two data triplets, namely ``source triplet'' and ``destination
triplet'' with the same anchor example, but sampled using different triplet
sampling strategies.
%
And the ``destination triplet'' is expected to result in a higher loss value
than another.
%
HM aims at creating adversarial examples of the ``source triplet'', that would
increase the corresponding loss value until that of the unperturbed
``destination triplet''.
%
This can also be interpreted as maximizing the loss value but ``stops early''
at a certain point.
%
The resulting perturbed ``source triplet'' can be used for adversarial training
of the model.
\subsection{Gradual Adversary}
\label{sec:32}

>>>>

Given an image triplet ($\mA$, $\mP$, $\mN$) sampled with a certain sampling
strategy (\eg, Random) within a mini-batch, we define its
\emph{hardness} as a scalar:
%
\begin{equation}
%
H(\mA,\mP,\mN)=d_\phi(\mA,\mP)-d_\phi(\mA,\mN),
%
\end{equation}
%
which is an internal part of the triplet loss.
%
For convenience we call this triplet as ``\emph{source} triplet'', and the
corresponding $H(\cdots)$ value as ``\emph{source} hardness'', denoted as
$H_\mathsf{S}$.

Then, \emph{Hardness Manipulation} (HM) aims to increase the \emph{source}
hardness $H_\mathsf{S}$ into a specified ``\emph{destination} hardness''
$H_\mathsf{D}$, by finding adversarial examples of the source triplet $(\mA,
\mP, \mN)$, \ie, $(\mA + {\vr}_a, \mP + {\vr}_p, \mN + {\vr}_n)$, where
$({\vr}_a, {\vr}_p, {\vr}_n)$ are the adversarial perturbations.
%
If we denote the hardness of the adversarially perturbed \emph{source} triplet
as $\tilde{H}_\mathsf{S}$, the HM can be implemented as follows:
%
\begin{equation}
	%
	\hat{\vr}_a, \hat{\vr}_p, \hat{\vr}_n = \argmin_{\vr_a,\vr_p,\vr_n}
	\big\|\max(0, H_\mathsf{D} - \tilde{H}_\mathsf{S} ) \big\|_2^2.
	%
	\label{eq:hm}
	%
\end{equation}
%
The $\max(0,\cdot)$ part in \cref{eq:hm} truncates the gradient when
$H_\mathsf{S}>H_\mathsf{D}$, automatically stopping the optimization, because
$H_\mathsf{S}$ is not desired to be reduced once being greater than
$H_\mathsf{D}$.
%
The \cref{eq:hm} is written in the L-$2$ norm form instead of the standard Mean Squared
Error because HM can be directly extended into vector form for a
mini-batch.
%
The optimization problem can be solved by Projected Gradient
Descent (PGD)~\cite{madry}.
%
And the resulting adversarial examples are used for adversarially training the
DML model:
%
$L_\text{T}(\phi(\mA+\hat{\vr}_a), \phi(\mP+\hat{\vr}_p),
\phi(\mN+\hat{\vr}_n))$.
%
We abbreviate the adversarial training with adversarial examples created through
this way as $\text{HM}[H_\mathsf{S},H_\mathsf{D}]$ in the following text.
%
And the overall procedure of HM is illustrated in \cref{fig:hm}.

@@

@@

Since the same $\Delta\vr$ can be used for both minimizing the HM objective
and maximizing the triplet loss, one potential advantage of HM is that
the gradients during the training process can be reused for creating
adversarial examples for much faster adversarial training, according to 
Free Adversarial Training~\cite{freeat}.
%
We leave this for future exploration.

% selection of destination hardness

\textbf{Destination Hardness} $H_\mathsf{D}$.
%
$\text{HM}[H_\mathsf{S},H_\mathsf{D}]$ is a very flexible method as various
\emph{destination} hardness $H_\mathsf{D}$ can be used, \eg, a constant, the
hardness of another benign triplet, or even a pseudo-hardness function.
%
The case of maximizing the triplet loss is equivalent to
$\text{HM}[H_\mathsf{S},2+\gamma]$, where $2+\gamma$ is the 
upperbound of triplet loss.

As pushing $\tilde{H}_\mathsf{S}$ towards the upperbound will easily lead to
model collapse, a valid $H_D\mathsf{D}$ should be choosed within the interval
$[H_\mathsf{S},2+\gamma]$ without rendering model collapse.
%
Thus, intuitively, $H_\mathsf{D}$ can be the hardness of another benign triplet
(with the same anchor) sampled with a strategy with a higher hardness
expectation, \ie, $E[H_\mathsf{D}] > E[H_\mathsf{S}]$.
%
Or at least the $\text{Var}[H_\mathsf{D}]$ of such sampling strategy has to be
large enough in order to create valid adversarial examples.
%
For instance, $H_\mathsf{D}$ can be the hardness of a Semihard~\cite{facenet}
triplet when the \emph{source} triplet is sampled with Random sampler.
%
Predictably, the model performance will be significantly influenced by the
triplet sampling strategies we chose for $H_\mathsf{D}$ in this case.
%
For convenience of further discussion, we denote the hardness of a Random
triplet, Semihard triplet, and Softhard triplet as $\mathcal{R}$,
$\mathcal{M}$, $\mathcal{S}$, respectively.

If we have a strong prior knowledge on what the \emph{destination} hardness
should be, we can even use a pseudo-hardness function $g(\cdot)$, which is a
customized scalar function.

\subsection{Gradual Adversary}
\label{sec:32}
